Longitudinal Principal Manifold Estimation

### 2023-10-17 paper/main.tex
overleaf updates

Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see\cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

At each time point, the estimated principal manifold has a spline form. As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point. This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
\begin{itemize}
    \item $\mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
    \item $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd;
    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$;
    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
\end{itemize}
This minimization is conducted by solving
\begin{equation}
  \left(
    \begin{array}{cc}
      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
      \boldsymbol{R} & \boldsymbol{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      \boldsymbol{s}_t \\
      \boldsymbol{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \boldsymbol{Y}_t \\
      \boldsymbol{0}
    \end{array}
  \right). \label{eq:13}
\end{equation}
The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
\begin{equation}
  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
\end{equation}

We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$. To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where with $\tau_t$ denoting the estimated mean squared distance of the PME model fit at time $t$ \meng{I would suggest writing the $\tau_t$ in a more precise way.}, the weights for each time are equal to 
$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$

With a given tuning value $\gamma$, we the fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$. This function takes the form

\begin{equation}
  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j, \label{eq:15}
\end{equation}
where, with matrix $\boldsymbol{A}$ defined as $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$, matrix $\boldsymbol{T}$ defined as $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to 
\begin{equation}
  \left(
  \begin{array}{ccc}
    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{A} + 2\gamma\boldsymbol{A} & 2\boldsymbol{A}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{T} \\
    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{A} & 2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{0} \\
    \boldsymbol{T}^{T} & \boldsymbol{0} & \boldsymbol{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \boldsymbol{\delta} \\
    \boldsymbol{\nu} \\
    \boldsymbol{m}
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{\Omega} \\
    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{\Omega} \\
    \boldsymbol{0}
  \end{array}
  \right), \label{eq:16}
\end{equation}
where $\boldsymbol{\Omega}=(\boldsymbol{\omega}_1,\ldots, \boldsymbol{\omega}_T)^\T$.

@@

\meng{From the methodology viewpoint, the fitting section is the most important part of this paper.}

\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction. The smoothness can be represented by the covariance kernel we use in the GPR. For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels. The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}

@@

Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows: \meng{What are the $Y_{i,t}$'s?}
\begin{equation}
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
\end{equation}
The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$. \meng{I would suggest some simulation studies showing the performance of the $\gamma^*$ something, like the Figure 4 of \cite{mengPrincipalManifoldEstimation2021}.}

Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands. \meng{Do some simulation studies and compare the leave-one-out CV and k-fold CV.}

@@

\meng{Make the notations in Algorithm \ref{alg:lpme} consistent with that in the description above.}

\meng{FYI, to fit a circle-like manifold, you may need to refer to the following papers: \cite{fasy2014confidence}, \cite{dunsonInferringManifoldsNoisy2022}.}

@@

@@

To assess the performance of the LPME algorithm, we compare its performance with that of two alternative approaches on simulated datasets. We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting. For each manifold, differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time were considered.

The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}. In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural changes in the underlying manifold over time. The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier. The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials. Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials. These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.

@@

In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run naively. The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}). In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen. Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$. In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.

Notably, the manifolds estimated in the simulated datasets included several self-intersecting manifolds, which methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit due to the self-consistency property discussed above. To avoid this issue, we augmented the simulated datasets with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition. Reaching a combined estimate after fitting PME to each partition individually requires "gluing" the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate. This source of error is avoided using the data augmentation approach. Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold. This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances. However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.

\begin{figure}%
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Augmenting the Unit Circle}
  \label{fig:unit_circle_augmentation}
\end{figure}

A factorial design is used to run the simulation studies, with the factor levels set as follows:
\begin{itemize}
  \item $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$
  \item Study duration: $\left\{1, 2, 5\right\}$
  \item Interval between images: $\left\{0.1, 0.25, 0.5\right\}$
  \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
\end{itemize}

The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times. All code required to reproduce the simulation results and the associated figures is provided at \texttt{https://github.com/rjzielinski/lpme-project}.

@@

\newpage

>>>>

Given that in most manifold learning tasks in practice data are often quite large, we begin by implementing an approach for selecting a small number number of points in $D$-dimensional space that represent the structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is assumed to be large. To reduce the sample size $I_t$, we use the HDMDE method (see \cite{mengPrincipalManifoldEstimation2021}, Section 5.3) to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the original time points to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with their corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$ that provide necessary information to characterize the shape of the original data. The number of clusters is estimated for each time point and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ from $I_t$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

At each time point, the estimated principal manifold has a spline form. As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point, with $N$ specified based on the maximum number of clusters identified in the data reduction step at any time point. This yields a set of outcomes, $\left\{Y_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $Y_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing the following function.
\begin{equation}\label{eq:splineL}
    \mathcal{L}(f_t) = (\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{Y}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t
\end{equation}    
where $\boldsymbol{Y}_t$ is the $N \times N$ matrix consisting of elements $\left\{Y_{i,t}\right\}_{i=1}^{N}$ at each time point $t$, the matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$, where $\eta_{d}(r) = r^{4 - d}\log r$ if $d$ is even, and $\eta_d(r) = r^{4-d}$ if $d$ is odd, the functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$; and the matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by $R_{ij} = \phi_i(\boldsymbol{r}_j)$.

By computing the derivative of $\mathcal{L}(f_t)$ in (\ref{eq:splineL}), we conduct the minimization of $\mathcal{L}(f_t)$ by solving the following system of equations. 
\begin{equation}
  \left(
    \begin{array}{cc}
      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
      \boldsymbol{R} & \boldsymbol{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      \boldsymbol{s}_t \\
      \boldsymbol{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \boldsymbol{Y}_t \\
      \boldsymbol{0}
    \end{array}
  \right). \label{eq:13}
\end{equation}
The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
\begin{equation}\nonumber
  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). 
\end{equation}

We denote the manifold coefficients at each time point $t$ as $\boldsymbol{\omega}_t = (\boldsymbol{s}_t^{*\T}, \boldsymbol{\alpha}_t^{*\T})^\T$. To limit the ability of a poorly-fitting PME estimate to greatly influence the results of the LPME model, we fit a weighted spline model to smooth over the manifold coefficients, where $\tau_t$ denote the estimated mean squared distance of the PME model fit at time $t$, the weights for each time are equal to 
$$w_t = \frac{1}{\tau_t\sum_{i=1}^{T}\frac{1}{\tau_i}}.$$

With a given tuning parameter value $\gamma$, we then fit a weighted cubic spline function using $\left\{t_i\right\}_{i=1}^T$ as predictors and $\left\{\mathbf{\omega}_t\right\}_t$ as the response values, with weights $\left\{w_t\right\}_t$ resulting in a function taking the form
\begin{equation}\nonumber
  g_{\gamma}(t) = \sum_{i=1}^{T}\delta_i \,\|t - t_i\|^{3} + \sum_{j=1}^{2}\nu_j\,\phi(t)_j,
\end{equation}
where by defining the $T \times T$ matrix $\boldsymbol{A}$ as defined  $\boldsymbol{A}_{ij} = \|t_i - t_j\|^{3}$ with $i,j = 1,\ldots, T$, the matrix $\boldsymbol{T}$ as defined $\boldsymbol{T}_{ij} = \phi_i(t_j)$, and $\boldsymbol{W} = \operatorname{diag}(w_1, \dots, w_T)$,  $\boldsymbol{\delta}=(\delta_1,\ldots,\delta_T)^\T$ and $\boldsymbol{\nu}=(\nu_1,\nu_2)^\T$ are the solutions to 
\begin{equation}
  \left(
  \begin{array}{ccc}
    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{A} + 2\gamma\boldsymbol{A} & 2\boldsymbol{A}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{T} \\
    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{A} & 2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{T} & \boldsymbol{0} \\
    \boldsymbol{T}^{T} & \boldsymbol{0} & \boldsymbol{0}
  \end{array}
  \right)\left(
  \begin{array}{c}
    \boldsymbol{\delta} \\
    \boldsymbol{\nu} \\
    \boldsymbol{m}
  \end{array}
  \right) = \left(
  \begin{array}{c}
    2\boldsymbol{A}\boldsymbol{W}\boldsymbol{\Omega} \\
    2\boldsymbol{T}^{T}\boldsymbol{W}\boldsymbol{\Omega} \\
    \boldsymbol{0}
  \end{array}
  \right), \label{eq:16}
\end{equation}
where $\boldsymbol{\Omega}=(\boldsymbol{\omega}_1,\ldots, \boldsymbol{\omega}_T)^\T$.

@@

@@

Once longitudinal estimates $F_{\gamma}(t, \boldsymbol{r})$ have been estimated for all provided tuning values, the optimal tuning parameter $\gamma^*$ is identified using leave-one-out cross validation. In this process, the coefficient smoothing function is computed while excluding all data and coefficients with time $t$ from consideration, for $t = 1, \dots, T$. We denote this function by $g_{\gamma}^{(t)}$. We can then assess the performance of the LPME model using tuning value $\gamma$ by averaging over estimates of MSD calculated using the original observations associated with the left-out time $t$ as follows.
\begin{equation}
  MSD(\gamma) = \frac{1}{T} \sum_{t=1}^{T}\frac{1}{I_t}\sum_{i=1}^{I_t}\|x_{i, t} - f_{\gamma}^{(t)}(t, \pi_{f_{\gamma}^{(t)}}(x_{i, t}))\|^2. \label{eq:18}
\end{equation}
The optimal tuning value $\gamma^*$ is then found to be the value of $\gamma$ that minimizes $MSD(\gamma)$. Note that alternative cross validation methods, such as $k$-fold cross validation, could easily be used in place of leave-one-out cross validation to better meet computational demands.

%\meng{Do some simulation studies and compare the leave-one-out CV and k-fold CV.}

@@

@@

\subsection{Self-intersecting Manifold Estimation}\label{ss:selfInt}

Notably, in many applications where manifold learning is implemented, the high-dimensional shapes of interest are self-intersecting manifolds. Methods based on the principal curve framework, including the principal manifold approach used here, are typically unable to fit such manifolds due to the self-consistency property. To avoid this issue, we propose a data augmentation approach with polar or spherical coordinates, depending on the value of $D$. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be considered as having $d = 1$ and $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

This approach to data augmentation offers an alternative to the method used in \cite{mengPrincipalManifoldEstimation2021} to fit the PME algorithm to self-intersecting manifolds, where the full dataset was manually partitioned to avoid intersections within each partition. Reaching a combined estimate after fitting PME to each partition individually requires ``gluing'' the estimated manifolds on each partition together, which may introduce error to the combined manifold estimate. This source of error is avoided using our proposed data augmentation approach. Additionally, scaling the added dimensions by a constant value enables the modification of the level of curvature present in the manifold. This may encourage improved performance when estimating manifolds that have too much curvature for PME to reach an appropriate fit under normal circumstances. However, an in depth comparison of these approaches is beyond the scope of this article and is left for future work.

\begin{figure}%
  \centering
  \subfloat[\centering Unaugmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D2}}}%
  \hfill
  \subfloat[\centering Augmented Unit Circle]{{\includegraphics[width=9cm]{unit_circle_D3}}}
  \caption{Illustration of our proposed data augmentation approach using the Unit Circle. The left panel (a) shows the depiction of the circle with $d=1$ and $D=2$, while the right panel (b) shows how our proposed use of considering this same manifold in $D=3$ by using the spherical coordinates results in a manifold that is not self-intersecting.}
  \label{fig:unit_circle_augmentation}
\end{figure}

@@

To assess the performance of the LPME algorithm, we compare its performance with that of two alternative approaches on simulated datasets. We consider settings in which $d = 1$ and $D = 2$, $d = 1$ and $D = 3$, and $d = 2$ and $D = 3$, with several manifolds being used to generate datasets in each setting. For each manifold, we consider differing values for the time duration, interval between observations, noise levels within and between time points, and the type and magnitude of structural changes in the underlying manifold over time. The embedding functions used to define each manifold are given in Table \ref{table:simulation_embeddings}. In these functions, the value of $\zeta$ and the functional form of $g(\cdot)$ are varied between trials and represent structural changes in the underlying manifold over time. The function $g(\cdot)$ may show change with respect to time that is constant, linear, quadratic, or sinusoidal, with $\zeta$ serving as a scalar multiplier. The values of $\iota$ represent within-image noise in the high-dimensional space, and are randomly sampled from a normal distribution with mean zero and variance that is varied between trials. Meanwhile, $\alpha$ and $\beta$ are vectors of length $d$ that are drawn from a normal distribution with mean equal to 1 and variance that is varied between trials. These values describe random fluctuations in the manifold between time points, reflecting noise introduced by different imaging sessions.

@@

In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point. The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package, developed using \texttt{R} (\cite{rSoftware2023}). In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen. Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$. In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.

A factorial design is used to run the simulation studies, with the factor levels set as follows:
\begin{itemize}
  \item $\alpha, \beta, \zeta \in \left\{0, 0.05, 0.1, 0.25, 0.5, 1.0\right\}$
  \item Study duration: $\left\{1, 2, 5\right\}$
  \item Interval between images: $\left\{0.1, 0.25, 0.5\right\}$
  \item Longitudinal change model: Constant, Linear, Quadratic, Sinusoidal
\end{itemize}
The sample size for each time point is set to 1,000 observations, so each embedding map is simulated a total of 7,776 times. All code required to reproduce the simulation results and the associated figures is provided at \href{https://github.com/rjzielinski/lpme-project}{\texttt{https://github.com/rjzielinski/lpme-project}}.

@@

\meng{FYI, we may also apply Gaussian process regression (GPR) to smooth the time direction. The smoothness can be represented by the covariance kernel we use in the GPR. For example, we may apply the Gaussian or \href{https://en.wikipedia.org/wiki/Matern_covariance_function}{Matérn} kernels. The taste of the GPR is similar to \cite{dunsonInferringManifoldsNoisy2022}, and we may cite \cite{dunsonInferringManifoldsNoisy2022}.}
\newpage
